---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Using the function read.csv and the function unz

```{r, load data}
data<-read.csv(unz("activity.zip","activity.csv"))
head(data)
```

Creating two variables to processing date and time informations
```{r, date and time variables}
date<-data$date
time<-data$interval
```

Now, we will adjust the variable time in order to be possible the convertion in class POSIXlt

```{r, adjust variable time}
time.adjusted<-sprintf("%04d", time)
head(time.adjusted)
```

Finally, the variable dateandtime is created. This variable will be used in all plots of time series
```{r, date and time variable}
dateandtime<-paste(date, time.adjusted)
dateandtime<-strptime(dateandtime, "%Y-%m-%d %H%M")
class(dateandtime)
head(dateandtime)
```

## What is mean total number of steps taken per day?

Firstly, we will calculate the total number of steps taken by day using the function tapply
```{r, steps per day}
stepspday<-tapply(data$steps,data$date,sum,na.rm=TRUE)
head(stepspday)
```

Now, we will create a histogram of the total steps taken each day.
```{r, histogram of steps taken in each day}
hist(stepspday,col="red", main="Total number of steps taken in each day",xlab="Total stpes")
```

Finally, we will  calculate the mean and median of total steps taken each day.
```{r, mean and median of steps per day}
options(scipen = 1, digits = 2)
stepspday.mean<-mean(stepspday[stepspday!=0])
stepspday.median<-median(stepspday[stepspday!=0])
```
So, the mean of steps taken each day is **`r stepspday.mean`** and the median of steps taken each day is **`r stepspday.median`**.

## What is the average daily activity pattern?

Firstly, we will calculate the mean of steps taken in each 5 minute interval. Again, we will use the function taplly.
```{r, steps per interval}
stepspinterval<-tapply(data$steps,data$interval,mean,na.rm=TRUE)
head(stepspinterval)
```

Now, we will plot the variable created above in function of each 5 minute interval.
```{r, plot of time series}

## The x axis variable dateandtime[1:288] was setted only for purpose of visualisation. The subset 1:288 is exactly one day of measured data.

plot(dateandtime[1:288],stepspinterval, type="l",ylab="Mean of steps taken in 5 minute interval", xlab="5 minute interval along 24 h", col="red", main="Average daily activity")
```

Finally, we will determine which 5 minute interval contains the biggest average of steps. 

For this, we will subset the vector of one day 5 minute interval for the maximum mean verified in stepspinterval funciotn.
```{r, interval with maximum average steps}
max.average<-dateandtime[1:288][stepspinterval==max(stepspinterval)]
```

So, the interval with biggest average steps is the 5 minute interval beggining in **`r max.average$hour`h:`r max.average$min`min**.

## Imputing missing values

This section begins with calculation of how many NAs is present in data provided.
```{r, how many NAs}
NA.total<-sum(is.na(data$steps))
```
The number os NAs in data provided is **`r NA.total`**.

Now, we will change the NA present in data for values. These values will be the mean verified for the same 5 minute interval.

Firstly, we will create another element in data. It will be the average steps for each interval, that was calculate in section above.
```{r, average steps in data}
average<-c(rep(stepspinterval,61))
data$average<-average
head(data)
```

Now, we will create another elemente in data. It will be steps without missing data. This element will be steps element if is not a missing number, otherwise it will be average element value.
```{r, steps with no missing value in data}
newsteps<-NULL
for (i in 1:length(data[,1])){
  if (is.na(data[i,]$steps)){
    newsteps[i]<-data[i,]$average
  } else {
    newsteps[i]<-data[i,]$steps
  }
}
data$newsteps<-newsteps
head(data)
```

Firstly, we will calculate the total number of steps (with no NAs) taken by day using the function tapply
```{r, steps per day with no NAs}
stepspday.no.NA<-tapply(data$newsteps,data$date,sum)
head(stepspday.no.NA)
```

Now, we will create a histogram of the variable newsteps.
```{r, histogram of steps without NAs}
hist(stepspday.no.NA,col="red", main="Total number of steps taken by each day (no NAs)",xlab="Total stpes")
```

Finally, we will  calculate the mean and median of total steps taken each day.
```{r, mean and median of steps per day (no NAs)}
options(scipen = 1, digits = 2)
stepspday.no.NA.mean<-mean(stepspday.no.NA)
stepspday.no.NA.median<-median(stepspday.no.NA)
```
So, the mean of steps taken each day is **`r stepspday.no.NA.mean`** and the median of steps taken each day is **`r stepspday.no.NA.median`**.

Only median has changed after treatment of missing values!!

## Are there differences in activity patterns between weekdays and weekends?

Firstly, we will create another element that will represent the of the week for each data collected.
```{r, weekday}
weekday<-NULL
for (i in 1:length(data[,1])){
  if (dateandtime[i]$wday>0 && dateandtime[i]$wday<6){
    weekday[i]<-"Weekday"
  } else {
    weekday[i]<-"Weekend"
  }
}
data$weekday<-weekday
head(data)
```

Now, we will split de data accordind to day of the week. The function used is split.
```{r, split weekday and weekend}
data.splited<-split(data,data$weekday)
str(data.splited)
```

Firstly, we will calculate the mean of steps taken in each 5 minute interval. Again, we will use the function taplly.
```{r, steps per interval in weekday and weekend}
stepspinterval.weekday<-tapply(data.splited$Weekday$newsteps,data.splited$Weekday$interval,mean)

stepspinterval.weekend<-tapply(data.splited$Weekend$newsteps,data.splited$Weekend$interval,mean)
head(stepspinterval.weekday)
head(stepspinterval.weekend)
```

Finally, it will be created the plot with average steps in each 5 minute interval for weekdays and weekends.
```{r, plot of time series in weekdays and weekends}
par(mfrow=c(2,1))

## The x axis variable dateandtime[1:288] was setted only for purpose of visualisation. The subset 1:288 is exactly one day of measured data.

plot(dateandtime[1:288],stepspinterval.weekday, type="l", col="red",main="Weekdays", ylim=c(0,300),xlab="",ylab="Average number of steps")

plot(dateandtime[1:288],stepspinterval.weekend, type="l",xlab="5 minute interval along 24 h", col="red", main="Weekend",ylim=c(0,300),ylab="Average number of steps taken")
```

